---
title: "R for Reproducible Scientific Analysis"
author: "Stacey Borrego"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview of Curriculum

### Part 1

* Introduction to R and RStudio
* Data Structures
* Exploring Data Frames
* Subsetting Data

### Part 2

* Control Flow
* Plotting with `ggplot2`

### Part 3

* Vectorization
* Functions
* Splitting and Combining Data Frames with `plyr`

### Part 4

* Data Frame Manipulation with `dplyr`
* Data Frame Manipulation with `tidyr`
* Producing Reports with `knitr`


# Part 1

### Introduction to R and RStudio

#### R as a calculator

```{r eval=FALSE}
3 + 5
12 / 7
3 * 4
8 ^ 2
log(8)
3 + 5 / 2
(3 + 5) / 2
``` 

#### Comparisons

```{r eval=FALSE}
1 == 1
1 != 2
1 < 2
1 <= 1
1 > 0
1 >= -9
``` 

#### Variables and Assigment

```{r eval=FALSE}
x <- 1/40
```

* Assignment operator (<-)
* Cannot start with numbers
* Case sensitive
* Avoid function names
* Avoid dots (.)
* Be consistent in the styling of your code
     + [R Style Guide](https://google.github.io/styleguide/Rguide.html)
     + Styles can include “lower_snake”, “UPPER_SNAKE”, “lowerCamelCase”, “UpperCamelCase”, etc
     + We will use "lower_snake" for readibility during this workshop 

Print the value of an object

```{r eval=FALSE}
x
```

#### Reassignment

```{r eval=FALSE}
x <- 100

x <- x + 1 
y <- x * 2

```

#### Vectorization

```{r eval=FALSE}
1:5
2^(1:5)
```

#### Managing Your Environment

```{r eval=FALSE}
ls()
rm()
sessionInfo()
?ls
```

* You can see what packages are installed by typing `installed.packages()`
* You can install packages by typing `install.packages("packagename")`, where packagename is the package name, in quotes.  
* You can update installed packages by typing `update.packages()`
* You can remove a package with `remove.packages("packagename")`
* You can make a package available for use with `library(packagename)`

#### Creating a New Project

**We’re going to create a new project in RStudio:**

* Click the “File” menu button, then “New Project”.
* Click “New Directory”.
* Click “New Project”.
* Type in the name of the directory to store your project, e.g. “my_project”.
* Click the “Create Project” button.


**Recommendations for project organization:**

* Put each project in its own directory, which is named after the project.
* Put text documents associated with the project in the doc directory.
* Put raw data and metadata in the data directory, and files generated during cleanup and  analysis in a results directory.
* Put source for the project’s scripts and programs in the src directory, and programs brought in from elsewhere or compiled locally in the bin directory.
* Name all files to reflect their content or function.

**Download the gapminder data from here.**

* Download the file (right mouse click on the link above -> “Save link as” / “Save file as”, or click on the link and after the page loads, press Ctrl+S or choose File -> “Save page as”)
* Make sure it’s saved under the name gapminder_data.csv
* Save the file in the data/ folder within your project.


### Data Structures



#### Data Types

```{r eval=FALSE}
typeof(3.14)
typeof(1L)
typeof(1+1i)
typeof('banana')
typeof(TRUE)
```

#### Vectors and Coersion

When R encounters a mix of types (here numeric and character) to be combined 
into a single vector, it will force them all to be the same type. 

The coercion rules go: logical -> integer -> numeric -> complex -> character, 
where -> can be read as are transformed into. 

```{r eval=FALSE}

combine <- c(2,6,3)
combine

quiz <- c(2,6,'3')
quiz

coercion <- c('a', TRUE)
coercion

coercion2 <- c(8, TRUE, FALSE)
coercion2

```

You can try to force coercion against this flow using the `as.` functions:

```{r eval=FALSE}
char <- c('0','2','4')

num <- as.numeric(char)
typeof(num)

log <- as.logical(num)
typeof(log)

```

Combine vectors

```{r eval=FALSE}
ab <- c('a', 'b')
ab

combine <- c(ab, 'SWC')
combine
```

Name elements in a vector

```{r eval=FALSE}
my_example <- 5:8
names(my_example)

names(my_example) <- c("a", "b", "c", "d")
names(my_example)
```

#### Reading and writing data

Make a toy dataset

```{r eval=FALSE}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                    weight = c(2.1, 5.0, 3.2),
                    likes_string = c(1, 0, 1))
```


Write it out as a csv file in your data directory

```{r eval=FALSE}
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```


Read the data file

```{r eval=FALSE}
cats <- read.csv(file = "data/feline-data.csv", stringsAsFactors = TRUE)
cats
```

Columns in data frames are vectors

```{r eval=FALSE}
str(cats$weight)
str(cats$likes_string)
str(cats$coat)
```

Types of data

```{r eval=FALSE}
typeof(cats$weight)

cats$likes_string
cats$likes_string <- as.logical(cats$likes_string)
cats$likes_string
```

Rename columns of a data frame

```{r eval=FALSE}
names(cats)
names(cats)[2] <- "weight_kg"
cats
```

#### Factors

```{r eval=FALSE}
coats <- c('tabby', 'tortoiseshell', 'tortoiseshell', 'black', 'tabby')
coats

str(coats)

CAT <- factor(coats)
class(CAT)
str(CAT)

typeof(coats)
typeof(CAT)

```

default factors are labelled in alphabetical order. You can change this by specifying the levels:

```{r eval=FALSE}
mydata <- c("case", "control", "control", "case")
factor_ordering_example <- factor(mydata, levels = c("control", "case"))
str(factor_ordering_example)
```

#### Lists

Another data structure you’ll want in your bag of tricks is the list. A list is simpler in some ways than the other types, because you can put anything you want in it. Remember everything in the vector must be the same basic data type, but a list can have different data types:

```{r eval=FALSE}
list_example <- list(1, "a", TRUE, 1+4i)
list_example

another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE )
another_list

typeof(cats)
```

We see that data.frames look like lists ‘under the hood’ - this is because a data.frame is really a list of vectors and factors, as they have to be - in order to hold those columns that are a mix of vectors and factors, the data.frame needs something a bit more flexible than a vector to put all the columns together into a familiar table. In other words, a data.frame is a special list in which all the vectors must have the same length.

In our cats example, we have an integer, a double and a logical variable. As we have seen already, each column of data.frame is a vector.


```{r eval=FALSE}
cats$coat
typeof(cats[,1])
str(cats[,1])
```

Each row is an observation of different variables, itself a data.frame, and thus can be composed of elements of different types.

```{r eval=FALSE}
cats[1,]
typeof(cats[1,])
str(cats[1,])
```

#### Matrices

```{r eval=FALSE}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example

class(matrix_example)
typeof(matrix_example)
str(matrix_example)
dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```

### Exploring Data Frames

See [Manipulating Data Frame](https://swcarpentry.github.io/r-novice-gapminder/05-data-structures-part2/index.html)

### Subsetting Data

See [Subsetting](https://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting/index.html)

# Part 2

### Control Flow

See [Control Flow](https://swcarpentry.github.io/r-novice-gapminder/07-control-flow/index.html)

### Plotting with `ggplot2`

See [ggplot2](https://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2/index.html)

# Part 3

### Vectorization
### Functions
### Splitting and Combining Data Frames with `plyr`

# Part 4

### Data Frame Manipulation with `dplyr`
### Data Frame Manipulation with `tidyr`
### Producing Reports with `knitr`


