---
title: "R for Reproducible Scientific Analysis"
author: "Stacey Borrego"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview of Curriculum

### Part 1

* Introduction to R and RStudio
* Data Structures
* Exploring Data Frames
* Subsetting Data

### Part 2

* Control Flow
* Plotting with `ggplot2`

### Part 3

* Vectorization
* Functions
* Splitting and Combining Data Frames with `plyr`

### Part 4

* Data Frame Manipulation with `dplyr`
* Data Frame Manipulation with `tidyr`
* Producing Reports with `knitr`


# Part 1 (Stacey is currently working on)

### Introduction to R and RStudio

#### R as a calculator

```{r eval=FALSE}
3 + 5
12 / 7
3 * 4
8 ^ 2
log(8)
3 + 5 / 2
(3 + 5) / 2
``` 

#### Comparisons

```{r eval=FALSE}
1 == 1
1 != 2
1 < 2
1 <= 1
1 > 0
1 >= -9
``` 

#### Variables and Assigment

```{r eval=FALSE}
x <- 1/40
```

* Assignment operator (<-)
* Cannot start with numbers
* Case sensitive
* Avoid function names
* Avoid dots (.)
* Be consistent in the styling of your code
     + [R Style Guide](https://google.github.io/styleguide/Rguide.html)
     + Styles can include “lower_snake”, “UPPER_SNAKE”, “lowerCamelCase”, “UpperCamelCase”, etc
     + We will use "lower_snake" for readibility during this workshop 

Print the value of an object

```{r eval=FALSE}
x
```

#### Reassignment

```{r eval=FALSE}
x <- 100

x <- x + 1 
y <- x * 2

```

#### Vectorization

```{r eval=FALSE}
1:5
2^(1:5)
```

#### Managing Your Environment

```{r eval=FALSE}
ls()
rm()
sessionInfo()
?ls
```

* You can see what packages are installed by typing `installed.packages()`
* You can install packages by typing `install.packages("packagename")`, where packagename is the package name, in quotes.  
* You can update installed packages by typing `update.packages()`
* You can remove a package with `remove.packages("packagename")`
* You can make a package available for use with `library(packagename)`

#### Creating a New Project

**We’re going to create a new project in RStudio:**

* Click the “File” menu button, then “New Project”.
* Click “New Directory”.
* Click “New Project”.
* Type in the name of the directory to store your project, e.g. “my_project”.
* Click the “Create Project” button.


**Recommendations for project organization:**

* Put each project in its own directory, which is named after the project.
* Put text documents associated with the project in the doc directory.
* Put raw data and metadata in the data directory, and files generated during cleanup and  analysis in a results directory.
* Put source for the project’s scripts and programs in the src directory, and programs brought in from elsewhere or compiled locally in the bin directory.
* Name all files to reflect their content or function.

**Download the gapminder data from here.**

* Download the file (right mouse click on the link above -> “Save link as” / “Save file as”, or click on the link and after the page loads, press Ctrl+S or choose File -> “Save page as”)
* Make sure it’s saved under the name gapminder_data.csv
* Save the file in the data/ folder within your project.


### Data Structures

#### Reading and writing data

Make a toy dataset

```{r eval=FALSE}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                    weight = c(2.1, 5.0, 3.2),
                    likes_string = c(1, 0, 1))
```


Write it out as a csv file in your data directory

```{r eval=FALSE}
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```

### Exploring Data Frames
### Subsetting Data

# Part 2

### Control Flow
### Plotting with `ggplot2`

# Part 3

### Vectorization
### Functions
### Splitting and Combining Data Frames with `plyr`

# Part 4

### Data Frame Manipulation with `dplyr`
### Data Frame Manipulation with `tidyr`
### Producing Reports with `knitr`


